---
title: "Association rules modeling"
output: rmarkdown::github_document
always_allow_html: yes
---

### Notebook objective
- Explore using association rules for online shopping data pattern/trend discovery.

### Association rules overview
- Aka Marketing Basket Analysis
- Historically, association rules models have been used to discover interesting patterns in retailer purchasing data. 
- However, association rules can be used in other domains like content streaming, social networks, etc. 
- This sort of analysis could power a similar products you might like feature, be used for selecting upsell products in a checkout flow, email marketing/promotions, etc.
- We generate rules for an item set of items. The item set could be one or more items. Think of an item here as a movie or retailer product SKU.
- Problem: brute force frequency calculation of all possible combinations of items is computationally expensive. 
- Modeling thresholds and [apriori principle](https://en.wikipedia.org/wiki/Apriori_algorithm) used to prune back the number of association calculations required.

### Key metrics: apriori model from arules package
- Support: frequency of users to buy A and B
- Confidence: of users to who buy A what percent also buy B
- Lift: how likely are users to buy B with A, while controlling for how popular B is. Positive value more likely. negative value less likely.
- Rules can we sorted/prioritized by various metrics depending on use case.
- Rules can included more more than two items (i.e (A, B => C).
- - [Arules intro paper](https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf)

###  Packages
```{r}
required_packages <- c('tidyverse', 'treemap', 'arules', 'arulesViz', 
                       'scales', 'readxl', 'janitor', 'mdsr')

for(p in required_packages) {
  library(p,character.only = TRUE)
}
```

### UCI Open Dataset
- [Online retail 2 dataset](http://archive.ics.uci.edu/ml/machine-learning-databases/00502/)

```{r}
df_2009_2010 <- read_excel("online_retail_II.xlsx", sheet = "Year 2009-2010")

df_2009_2010_tbl <- tibble::as_tibble(
    df_2009_2010, 
    .name_repair = janitor::make_clean_names
  ) %>%
  mutate(data_file_key = "2009_2010")

df_2010_2011 <- read_excel("online_retail_II.xlsx", sheet = "Year 2010-2011")

df_2010_2011_tbl <- tibble::as_tibble(
    df_2010_2011, 
    .name_repair = janitor::make_clean_names
  ) %>% 
  mutate(data_file_key = "2010_2011")

combined_years_df <- bind_rows(df_2009_2010_tbl, df_2010_2011_tbl)

combined_years_df <- combined_years_df %>%
  mutate(canceled_order = !is.na(str_extract(invoice, "^C"))) %>%
  filter(canceled_order==F & !is.na(customer_id)) %>%
  # filter out adjustment records
  filter(!str_detect(stock_code, "ADJUST"))

### clean product descriptions of punctuation
### prevents parsing issues downstream when using ":" to separate products
combined_years_df <- combined_years_df %>%
  mutate(description = str_replace_all(description, "[:punct:]", ""),
         description = str_replace_all(description, "  ", " "))
```

### Some product codes have multiple descriptions
- need consistent product descriptions for arules generation
- ~12% of product codes have more than one description

```{r}
### number of items with multiple descriptions
combined_years_df %>%
  group_by(stock_code) %>%
  summarise(unique_description_count = n_distinct(description)) %>%
  count(unique_description_count) %>%
  mutate(percent_total = n/sum(n))

multiple_description_codes <- combined_years_df %>%
  group_by(stock_code) %>%
  summarise(unique_description_count = n_distinct(description)) %>%
  filter(unique_description_count>1)
```

### Get the last description for product codes with multiple descriptions
```{r}
descriptions_to_use <- combined_years_df %>%
  filter(stock_code %in% unique(multiple_description_codes$stock_code)) %>%
  group_by(stock_code) %>%
  summarise(last_description = last(description, order_by = invoice_date)) %>%
  ungroup()

### replace product descriptions for codes with multiples
combined_years_df <- combined_years_df %>%
  left_join(descriptions_to_use) %>%
  mutate(description = ifelse(
    is.na(last_description), 
    description, 
    last_description
    )
  ) %>%
  select(-last_description)

### check that all product codes now have one unique description
combined_years_df %>%
  group_by(stock_code) %>%
  summarise(unique_descriptions = n_distinct(description)) %>%
  count(unique_descriptions)
```

### Typical invoice size and quanity
```{r}
### invoice level summary starts
combined_years_df %>%
  group_by(invoice) %>%
  summarise(products_purchased = n_distinct(stock_code),
            average_quanity = mean(quantity)) %>%
  mdsr::skim() %>%
  select(-n)
```

### Customer level summary stats
```{r}
combined_years_df %>%
  group_by(cust_char = as.character(customer_id)) %>%
  summarise(number_of_invoices = n_distinct(invoice),
            products_purchased = n_distinct(stock_code)) %>%
  mdsr::skim() %>%
  select(-n)
```

### Generate transaction format df (one row per invoice)
- for each invoice, we concat the products purchased together separated by ":"

```{r}
transaction_format <- combined_years_df %>%
  group_by(invoice) %>%
  summarise(products_purchased = paste0(unique(description), collapse = ":")) %>%
  select(-invoice) %>%
  ungroup()

write_csv(transaction_format, "online_retail_2_transactions.csv")

sample_n(transaction_format, 5) %>%
  glimpse()
```

### Read in data using arules package format
- prevent duplicates from getting removed 
- input data is in basket format with one row
- representing an invoice of products

```{r}
### input data is in basket format with one row
### representing an invoice of products
trans <- arules::read.transactions("online_retail_2_transactions.csv",
                           format = "basket",
                           sep = ":")
```

### Item Freq
```{r}
item_freq_df <- tibble(item = names(itemFrequency(trans)),
                    item_order_frequency = itemFrequency(trans))

item_freq_df %>%
  arrange(-item_order_frequency) %>%
  head(25) %>%
  ggplot(aes(y=reorder(item, item_order_frequency),
             x=item_order_frequency)) +
  geom_col() +
  theme(axis.text.y = element_text(size=6)) +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 24))
```

```{r}
support_set_1 <- eclat(trans,
       parameter = list(
             support=0.01,
             minlen=2,
             maxlen=2
       ))

support_set_1 %>%
  inspect()
  
  mutate(items = str_remove_all(items, "\\{"),
         items = str_remove_all(items, "\\}")) %>%
  separate(items, c("a", "b"), sep = ",")

view(support_df_l1)
```

```{r}
retail_rules <- apriori(trans,
                        parameter = list(
                          support=0.01,
                          confidence=0.01,
                          minlen=2,
                          maxlen=4
                          )
                        )

summary(retail_rules)

retail_rules %>% 
      sort(by="support") %>%
      head(10) %>%
      inspect() %>%
      view()

### invoices by customer
combined_years_df %>%
  group_by(customer_id) %>%
  summarise(invoice_count = n_distinct(invoice),
            total_quantity = sum(quantity)) %>%
  view()
```


```{r}
library(ggraph)
library(igraph)

product_support_set_df <- retail_rules %>% 
      sort(by="support") %>%
      head(50) %>%
      inspect()

graph_set <- product_support_set_df %>%
  select(lhs, rhs, support) %>%
   graph_from_data_frame()

ggraph(graph_set) + 
    geom_edge_link0(alpha=0.4, color="grey40", show.legend = F) + 
    geom_node_point() +
    geom_node_label(aes(label = name), 
                    color="dodgerblue",
                    size=2, repel = TRUE, show.legend = F,
                    segment.color="dodgerblue") +
  labs(title="Top 50 Most Common Item Purchase Pairs")
```


### ggraph exploration products
```{r}
library(ggraph)
library(igraph)

### top x most common level 1 pairs
graph_df_2 <- product_support_set_df %>% arrange(-support) %>% head(200)

graph_2 <- graph_from_data_frame(graph_df_2)

# Not specifying the layout - defaults to "auto"
ggraph(graph_2, layout = "stress") + 
    geom_edge_link0(alpha=0.4, color="grey40", show.legend = F) + 
    geom_node_point() +
    geom_node_label(aes(label = name), 
                    color="dodgerblue",
                    size=2, repel = TRUE, show.legend = F,
                    segment.color="dodgerblue") +
  labs(title="Top 200 Most Common Item Purchase Pairs")



### Open Questions
- why would the customer id be NULL?




