---
title: "Association rules modeling"
output: rmarkdown::github_document
always_allow_html: yes
---

### Notebook objective
- Explore using association rules for online shopping data pattern/trend discovery.

### Association rules overview
- Aka Marketing Basket Analysis
- Historically, association rules models have been used to discover interesting patterns in retailer purchasing data. 
- However, association rules can be used in other domains like content streaming, social networks, etc. 
- This sort of analysis could power a similar products you might like feature, be used for selecting upsell products in a checkout flow, email marketing/promotions, etc.
- We generate rules for an item set of items. The item set could be one or more items. Think of an item here as a movie or retailer product SKU.
- Problem: brute force frequency calculation of all possible combinations of items is computationally expensive. 
- Modeling thresholds and [apriori principle](https://en.wikipedia.org/wiki/Apriori_algorithm) used to prune back the number of association calculations required.

### Key metrics: apriori model from arules package
- Support: frequency of users to buy A and B
- Confidence: of users to who buy A what percent also buy B
- Lift: how likely are users to buy B with A, while controlling for how popular B is. Positive value more likely. negative value less likely.
- Rules can we sorted/prioritized by various metrics depending on use case.
- Rules can included more more than two items (i.e (A, B => C).
- - [Arules intro paper](https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf)

###  Packages
```{r}
required_packages <- c('tidyverse', 'treemap', 'arules', 'arulesViz', 
                       'scales', 'readxl', 'janitor', 'mdsr')

for(p in required_packages) {
  library(p,character.only = TRUE)
}
```


```{r}
df_2009_2010 <- read_excel("online_retail_II.xlsx", sheet = "Year 2009-2010")

df_2009_2010_tbl <- tibble::as_tibble(
    df_2009_2010, 
    .name_repair = janitor::make_clean_names
  ) %>%
  mutate(data_file_key = "2009_2010")

df_2010_2011 <- read_excel("online_retail_II.xlsx", sheet = "Year 2010-2011")

df_2010_2011_tbl <- tibble::as_tibble(
    df_2010_2011, 
    .name_repair = janitor::make_clean_names
  ) %>% 
  mutate(data_file_key = "2010_2011")

combined_years_df <- bind_rows(df_2009_2010_tbl, df_2010_2011_tbl)

combined_years_df <- combined_years_df %>%
  mutate(canceled_order = !is.na(str_extract(invoice, "^C"))) %>%
  filter(canceled_order==F & !is.na(customer_id)) %>%
  # filter out adjustment records
  filter(!str_detect(stock_code, "ADJUST"))

### clean product descriptions of punctuation
### prevents parsing issues downstream when using ":" to separate products
combined_years_df <- combined_years_df %>%
  mutate(description = str_replace_all(description, "[:punct:]", ""),
         description = str_replace_all(description, "  ", " "))
```

### Some product codes have multiple descriptions
- need consistent product descriptions for arules generation
- ~12% of product codes have more than one description

```{r}
### number of items with multiple descriptions
combined_years_df %>%
  group_by(stock_code) %>%
  summarise(unique_description_count = n_distinct(description)) %>%
  count(unique_description_count) %>%
  mutate(percent_total = n/sum(n))

multiple_description_codes <- combined_years_df %>%
  group_by(stock_code) %>%
  summarise(unique_description_count = n_distinct(description)) %>%
  filter(unique_description_count>1)
```
### Get the 

```{r}
combined_years_df %>%
  filter(stock_code %in% multiple_description_codes %>% pull(stock_code)) %>%
  group_by(stock_code) %>%
  summarise(last_description = last(description, order_by = -invoice_date))
```




### Typical invoice size and quanity
```{r}
### invoice level summary starts
combined_years_df %>%
  group_by(invoice) %>%
  summarise(products_purchased = n_distinct(stock_code),
            average_quanity = mean(quantity)) %>%
  mdsr::skim() %>%
  select(-n)
```

### Customer level summary stats
```{r}
combined_years_df %>%
  group_by(cust_char = as.character(customer_id)) %>%
  summarise(number_of_invoices = n_distinct(invoice),
            products_purchased = n_distinct(stock_code)) %>%
  mdsr::skim() %>%
  select(-n)
```

### Generate transaction format df | one row per invoice
```{r}
glimpse(combined_years_df)

combined_years_df %>%
  summarise(unique_descriptions = n_distinct(description),
            unique_stock_codes = n_distinct(stock_code))

### find instances where the same product description occurs multiple times in an invoice
combined_years_df %>%
  group_by(invoice, description) %>%
  mutate(description_record_count = n()) %>%
  ungroup() %>%
  filter(description_record_count>1) %>%
  view()

combined_years_df %>%
  group_by(stock_code) %>%
  summarise(description_count = n_distinct(description)) %>%
  ungroup() %>%
  filter(description_count>1) %>%
  view()

combined_years_df %>%
  filter(str_detect(stock_code, "ADJ")) %>%
  view()

# combined_years_df %>%
#   filter(invoice==489488) %>% 
#   filter(description=="CHARLIE LOLA BLUE HOT WATER BOTTLE")
#   group_by(invoice) %>%
#   summarise(unique_products = unique(description))

transaction_format <- combined_years_df %>%
  group_by(invoice) %>%
  ### paste on unique prevents duplicate product descriptions showing up in output
  summarise(products_purchased = paste0(unique(description), collapse = ":")) %>%
  select(-invoice) %>%
  ungroup()

write_csv(transaction_format, "online_retail_2_transactions.csv")

sample_n(transaction_format, 5) %>%
  glimpse()
```

### Read in data using arules package format
- prevent duplicates from getting removed 


```{r}
### input data is in basket format with one row
### representing an invoice of products
trans <- arules::read.transactions("online_retail_2_transactions.csv",
                           format = "basket",
                           sep = ":",
                           # cleans
                           rm.duplicates=TRUE)
# summary(trans)
```
### Item Freq
```{r}
item_freq_df <- tibble(item = names(itemFrequency(trans)),
                    item_order_frequency = itemFrequency(trans))

item_freq_df %>%
  arrange(-item_order_frequency) %>%
  head()

support_set_1 <- eclat(trans,
       parameter = list(
             support=0.01,
             minlen=2,
             maxlen=2
       ))

support_set_1 %>%
  inspect()
  
  mutate(items = str_remove_all(items, "\\{"),
         items = str_remove_all(items, "\\}")) %>%
  separate(items, c("a", "b"), sep = ",")

view(support_df_l1)
```




### Open Questions
- why would the customer id be NULL?




