---
title: "Association rules modeling | UCI retail shopping 2 dataset"
output: rmarkdown::github_document
always_allow_html: yes
---

```{r echo=FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

### Notebook objective
- Explore using association rules for e-commerce pattern/trend discovery.

### Association rules overview (AKA marketing basket analysis)
- Historically, association rules models have been used to discover interesting patterns in retailer/shopping purchasing data. 
- However, association rules can be used in other domains like content streaming, social networks, etc. 
- This technique could power a similar products you might like feature, be used for electing upsell products in a checkout flow, email marketing/promotions, etc.
- With association rules we generate rules for an item set of items. The item set could be one or more items. Think of an item here as a movie or retailer product SKU.
- Brute force frequency calculation of all item set combinations is computationally expensive. 
- Model thresholds and [apriori principle](https://en.wikipedia.org/wiki/Apriori_algorithm) are used to prune back the number of association calculations required. 

### Key metrics
- Support: frequency of users to purchase an item (e.g. could be just product A or product A & B, etc)
- Confidence: of users to who buy A what percent also buy B
- Lift: how likely are users to buy B with A, while controlling for how popular B is (positive value more likely, negative value less likely).
- Rules can we sorted/prioritized by various metrics depending on use case.
- Rules can included more more than two items (i.e (A, B => C).
- [Arules Package Intro Paper](https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf)

###  Packages
```{r}
required_packages <- c('tidyverse', 'arules', 'arulesViz', 
                       'scales', 'readxl', 'janitor')

for(p in required_packages) {
  library(p,character.only = TRUE)
}
```

### UCI Open Dataset
- [Data dictionary and background context](https://archive.ics.uci.edu/ml/datasets/Online+Retail+II)
- [Online retail 2 dataset](http://archive.ics.uci.edu/ml/machine-learning-databases/00502/)

```{r}
# read online retail Execl from working directory
df_2010_2011_raw <- read_excel("online_retail_II.xlsx", sheet = "Year 2010-2011")

# clean column names
df_2010_2011 <- 
  tibble::as_tibble(
    df_2010_2011_raw, 
    .name_repair = janitor::make_clean_names
  )

# filter out adjustment records, cancel records, 
df_2010_2011 <- df_2010_2011 %>%
  mutate(canceled_order = !is.na(str_extract(invoice, "^C"))) %>%
  filter(canceled_order == F & !is.na(customer_id)) %>%
  filter(!str_detect(stock_code, "ADJUST"))

### clean product descriptions of punctuation
### prevents parsing issues downstream when using ":" for data setup
df_2010_2011 <- df_2010_2011 %>%
  mutate(description = str_replace_all(description, "[:punct:]", ""),
         description = str_replace_all(description, "  ", " "))
```

### Some product codes have multiple descriptions
- Need consistent product descriptions for arules generation.
- ~5% of product codes have more than one description.

```{r}
### number of items with multiple descriptions
df_2010_2011 %>%
  group_by(stock_code) %>%
  summarise(unique_description_count = n_distinct(description)) %>%
  group_by(unique_description_count) %>%
  summarise(stock_code_count = n_distinct(stock_code)) %>%
  mutate(percent_total = stock_code_count/sum(stock_code_count)*100)

multiple_description_codes <- df_2010_2011 %>%
  group_by(stock_code) %>%
  summarise(unique_description_count = n_distinct(description)) %>%
  filter(unique_description_count>1)
```

### Product description clean up
- Data cleaning approach to handle cases where stock codes could have multiple descriptions.

```{r}
descriptions_to_use <- df_2010_2011 %>%
  filter(stock_code %in% unique(multiple_description_codes$stock_code)) %>%
  group_by(stock_code) %>%
  summarise(last_description = last(description, order_by = invoice_date)) %>%
  ungroup()

### replace product descriptions for codes with multiples
df_2010_2011 <- df_2010_2011 %>%
  left_join(descriptions_to_use) %>%
  mutate(description = ifelse(
    is.na(last_description), 
    description, 
    last_description
    )
  ) %>%
  dplyr::select(-last_description)

### check that all product codes now have one unique description
df_2010_2011 %>%
  group_by(stock_code) %>%
  summarise(unique_descriptions = n_distinct(description)) %>%
  dplyr::count(unique_descriptions)
```

### Typical invoice size and quanity
- Typical invoice order contains 15 (median) stock codes.

```{r}
### invoice level summary starts
df_2010_2011 %>%
  group_by(invoice) %>%
  summarise(products_purchased = n_distinct(stock_code),
            average_quanity = mean(quantity),
            .groups = "drop") %>%
  summary()
```

### Customer level summary stats
- Median number of invoices per customer: 2.

```{r}
df_2010_2011 %>%
  group_by(cust_char = as.character(customer_id)) %>%
  summarise(number_of_invoices = n_distinct(invoice),
            products_purchased = n_distinct(stock_code),
            .groups = "drop") %>%
  summary()
```

### Generate transaction format (one row per invoice)
- For each invoice, we concat the products purchased by ":".
- This approach generates a basket format csv which is a input format for the arules package.

```{r}
transaction_format <- df_2010_2011 %>%
  group_by(invoice) %>%
  summarise(products_purchased = paste0(unique(description), collapse = ":")) %>%
  dplyr::select(-invoice) %>%
  ungroup()

write_csv(transaction_format, "online_retail_2_transactions.csv")
```

### Read in data using arules package format
- Input data is in basket format with one row representing a character string of products included in a customers invoice.

```{r}
trans <- arules::read.transactions("online_retail_2_transactions.csv",
                           format = "basket",
                           sep = ":")
```

### Most Commons Products
- Popular products related to party supplies and party gifts (aligns with data source info which states the data is from a company which mainly sells unique all-occasion gift-ware).

```{r}
item_freq_df <- tibble(item = names(itemFrequency(trans)),
                    item_order_frequency = itemFrequency(trans))

item_freq_df %>%
  arrange(-item_order_frequency) %>%
  head(25) %>%
  ggplot(aes(y=reorder(item, item_order_frequency),
             x=item_order_frequency)) +
  geom_col() +
  theme(axis.text.y = element_text(size=6)) +
  scale_y_discrete(label = function(x) stringr::str_trunc(x, 24)) +
  scale_x_continuous(labels = scales::percent_format(accuracy=1),
                     breaks = seq(0, 1, by=0.01)) +
  theme_minimal() +
  labs(title="Top 25 Most Common Products Purchased",
       y="Product",
       x="Support (Purchase Frequency)")
```

### Most common product pairs ordered
- Eclat algorithm to find most common item pairs, etc.

```{r}
item_pairs <- eclat(trans, parameter = list(support=0.005, minlen=2, maxlen=2))

sort(item_pairs, by = 'support') %>% 
  as("data.frame") %>%
  head(10) %>%
  dplyr::select(-transIdenticalToItemsets)
```
  
##### Exclude item pairs that contain BAG or BOX
```{r}
sort(item_pairs, by = 'support') %>%
  as("data.frame") %>%
  filter(!str_detect(items, "BAG | BOX")) %>%
  head(10) %>%
  dplyr::select(-transIdenticalToItemsets)
```

### Most Commons Product Trios Ordered
```{r}
item_trios <- eclat(trans, parameter = list(support=0.005, minlen=3, maxlen=3))

sort(item_trios, by = 'support') %>%
  as("data.frame") %>%
  head(10) %>%
  dplyr::select(-transIdenticalToItemsets)
```

### Apriori algorithm
- Eclat generates metrics per itemset `{A, B}`.
- Apriori algo generates metrics comparing itemsets and outputs rules `{A => B}`.

```{r}
retail_rules <- apriori(trans, 
                        parameter = list(
                          support=0.005,
                          confidence=0.005,
                          minlen=2,
                          maxlen=4
                          )
                        )

summary(retail_rules)
```

##### Top 10 rules by support
```{r}
retail_rules %>% 
  as("data.frame") %>%
  arrange(-support) %>%
  head(10)
```
##### Top 10 rules by confidence
```{r}
retail_rules %>% 
  as("data.frame") %>%
  arrange(-confidence) %>%
  head(10)
```

##### Top 10 rules by lift
```{r}
retail_rules %>% 
  as("data.frame") %>%
  arrange(-lift) %>%
  head(10)
```

### Visualizing rules
- Plot function part of the arulesViz package is used to visualize rules. [Documentation on customization options](https://rdrr.io/cran/arulesViz/man/plot.html)
- Visualizations and rule metrics to focus on likely to be informed by business objective.

##### Scatter plot of rule support, confidence, lift metrics
- Top 100 rules ranked by support.

```{r}
plot(sort(retail_rules, by="support")[1:100])
```

##### Network graph of rules

Top 20 rules ranked by support.
```{r}
plot(sort(retail_rules, by="support")[1:20], 
     method="graph", 
     cex=0.7)
```

Top 20 rules ranked by confidence. Adjust plot params.
```{r}
plot(sort(retail_rules, by="confidence")[1:20], 
     method="graph", 
     measure="confidence",
     shading = "support",
     cex=0.7)
```

### Ideas for applying these techniques in an actionable business setting 
- Marketing teams could build packages/promos for common products customers buy together (using above data).
- Have a UX flow that makes it easy to buy variations of parent SKU (i.e. a few clicks to buy 200 lunch boxes with 10 different color patterns). Arules data can suggest product areas to focus on first.
- Additional product meta data could be used to explore purchasing trends at the category or sub category level.
- Segment data by customer type and purchase frequency (i.e. geo, wholesalers, SMBs, individual customers).
- Recommendation features that leverage arules package rules to suggest products a customer might also like.
- Interactive data visualizations could be used to empower internal stakeholders to explore arules trends via self-service tools.
